---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(corrplot)
library(dplyr)
library(knitr)
library(readr)
library(here)
library(kableExtra)

merged_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
show_col_types = FALSE
```


# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}
```{r plot-chunk, fig.cap="Correlation matrix of numeric variables", fig.width=8, fig.height=7, echo=FALSE}
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Assuming 'lg_rank' is your league rank variable
merged_data$transformed_rank <- max(merged_data$lg_rank) + 1 - merged_data$lg_rank
# Calculate the correlation matrix for numeric columns
cor_matrix <- cor(merged_data |> select(where(is.numeric)), use = "complete.obs")

corrplot(cor_matrix, method = "circle", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, tl.cex = 0.7)

# Manually add title
mtext("Correlation Matrix for Toronto FC's Performance Features", side=3, line=3, cex=1.5)
```



# Model


## Model set-up


### Model justification


# Results

# Discussion

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

\newpage

\appendix

# Appendix {-}

```{r}
#| label: tbl-lg-rank-correlation
#| tbl-cap: Correlation of Variables with lg_rank
#| echo: false
#| warning: false

# Convert the correlation matrix to a long-format dataframe, focusing on lg_rank
cor_df_long_lg_rank <- cor_matrix |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  pivot_longer(cols = -Variable, names_to = "Related_Variable", values_to = "Correlation") |>
  filter(Variable == "lg_rank" ) |>
  distinct()

# Use knitr::kable to display the table, focusing on correlations with lg_rank
knitr::kable(cor_df_long_lg_rank, padding = 5, format = "latex", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = c("striped", "scale_down"))
```


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

## Diagnostics

\newpage


# References


